Multidimensional signals and analytic flexibility: Estimating degrees of freedom in human speech analyses

### 2021-09-12 RR_manuscript/Draft_RR.Rmd
final edits

Speech is a complex signal that is characterized by structurally different acoustic landmarks distributed throughout different temporal domains.
Thus, choosing how to measure a phenomenon of interest is an important analytic step.
Take for example the following sentence in (1).

@@

The multiplicity of phonetic measurements grows exponentially if we look at larger temporal domains as is the case for suprasegmental aspects of speech.
For example, studies investigating acoustic correlates of word stress (e.g. the difference between *ínsight* and *incíte*) use a wide variety of measurements, including temporal characteristics (duration of certain segments or sub-segmental intervals), spectral characteristics (intensity, formants, and spectral tilt), and measurements related to fundamental frequency (f0) [e.g., @gordon2017acoustic].
Moving on to the expression of higher-level functions, like information structure and discourse pragmatics, relevant acoustic cues can be distributed throughout even larger domains, such as phrases and whole utterances [e.g., @ladd2008intonational].
Differences in position, shape, and alignment of pitch modulations over multiple locations within a sentence are correlated with differences in discourse functions [e.g., @niebuhr2011].
The latter can also be expressed by global vs local pitch modulations [@heuven2002], as well as acoustic information within the temporal or spectral domain [e.g., @van2005speech].
Extra-linguistic information, like the speaker's intentions, levels of emotional arousal or social identity, are also conveyed by broad-domain parameters, such as voice quality, rhythm, and pitch [@foulkes2006; @ogden2004; @white2009].

In short, when testing hypotheses on speech production data, researchers are faced with many choices and possibilities.
The larger the functional domain (e.g. segments vs words vs utterances), the higher the number of conceivable operationalizations.
For example, several decisions have to be made when comparing the two realizations of the sentence in (1), one of which is intended to signal emphasis on *another* and one of which emphasizes *Zoom*.

@@

Do we compare only the word *another* in (2a) and (2b), or also the word *Zoom*?
Do we measure utterance-wide acoustic profiles, whole words, or just the stressed syllables?
Do we average across the chosen time domain or do we measure a specific point in time?
Do we measure fundamental frequency, intensity, or something else [@stevens2000]?

@@

When looking at phrase-level temporal domains, the number of alternative analytic pipelines increases substantially.
Figure \@ref(fig:forkingPaths)A shows a typical example of a decision tree with which speech researchers are often confronted.
Each of the four analytic decisions in the example have different possible options.
Here only one particular path has been taken, and a different one would likely produce quite different results.
Once we have decided to compare fundamental frequency of the word *another* across the two utterances, there are still many choices to be made, all of which need to be justified.
As Figures \@ref(fig:forkingPaths)B-C illustrate, we could measure f0 at specific points in time like the onset of the temporal window, the offset, or the midpoint.
We could also measure the value or time of the minimum or maximum f0 value.
We could summarize f0 across the entire window and extract the mean, median or standard deviation of f0, all of which have been used to analyze speech data in previous work [see @gordon2017acoustic].
But the journey in the garden of analytic paths goes on.
Other important operationalization steps could involve filtering the audio signal, smoothing the extracted f0 track, removing values that substantially deviate from surrounding values or expectations, either manually or automatically, and so on.

These decisions are intended to be made prior to any statistical analysis, but are at times revised a posteriori in light of unforeseen or surprising outcomes (i.e. after data collection and/or preliminary analyses).
These myriads of possible decisions are multiplied by those researcher degrees of freedom related to statistical analysis [e.g. @wicherts2016].
In sum, speech data is made of complex physical signals that generates an as-of-yet unappreciated amount of analytic flexibility in the choice of measures and operationalizations.
The present paper probes this garden of forking paths in the analysis of speech.
To assess the variability in data analysis pipelines, including both operationalization and statistical analysis, across independent researchers, we provided analytic teams with an experimentally elicited speech production data set.
The data set derives from the unpublished research project *Prosodic encoding of redundant referring expressions*, which set out to investigate whether speakers acoustically modify utterances to signal unexpected referring expressions.^[Results of this research project were neither published nor publicly presented and are stored on a private OSF repository.] 
In the following section we introduce the research question and the experimental procedure of said project, and we describe the resulting data set as used in the current study.

@@

Referring is one of the most basic and prevalent uses of language and one of the most widely researched areas in language science.
It is an open question how speakers choose a referring expression when they want to refer to a specific entity like a banana.
The context within which an entity occurs (i.e., with other non-fruits, other fruits, or other bananas) plays a large part in determining the choice of referring expressions.
Generally, speakers aim to be as informative as possible to uniquely establish reference to the intended object, but they are also resource-efficient in that they avoid redundancy [@grice1975logic].
Thus one would expect the use of a modifier, for example, only if it is necessary for disambiguation.
For instance, one might use the adjective *yellow* to describe a banana in a situation in which there is a yellow and a less ripe green banana available, but not when there is only one banana to begin with.

Despite the coherent idea that speakers are both rational and efficient, there is much evidence that speakers are often over-informative.
Speakers use referring expressions that are more specific than strictly necessary for the unambiguous identification of the intended referent [@sedivy2003pragmatic; @westerbeek2015stored; @rubio2016redundant], which has been argued to facilitate object identification and make communication between speakers and listeners more efficient [@arts2011overspecification; @paraboni2007generating; @rubio2016redundant].
Recent findings suggest that the utility of a referring expression depends on how good it is for a listener (compared to other referring expressions) to identify a target object.
For example, @degen2020redundancy showed that modifiers that are less typical for a given referent (e.g. a blue banana) are more likely to be used in an over-informative scenario (e.g. when there is just one banana).
This account, however, has mainly focused on content selection [@gatt2013we].

Even when morphosyntactically identical expressions are involved, speakers can modulate utterances via suprasegmental acoustic properties like temporal and spectral modifications [e.g., @ladd2008intonational].
Most prominently, languages use intonation to signal discourse relationships between referents, among other functions.
Intonation marks discourse-relevant referents for being new or given information, to guide the listeners’ interpretation of incoming messages.
Beyond structuring information relative to the discourse, a few studies suggested that speakers might use intonation to signal atypical lexical combinations [e.g. @dimitrova2008prosodic; @dimitrova2009did].
Referential expressions such as *blue banana* were produced with greater prosodic prominence than more typical referents such as *yellow banana*.
These results are in line with the idea of resource-efficient, rational language users who modulate their speech in order to facilitate listeners' comprehension.
However, the above studies are based on a rather small sample size (10 participants) and on potentially anti-conservative statistical analyses, leaving reason to doubt the generalizability of the studies' conclusions.

To further illuminate the question of whether speakers modify speech to signal atypical referents, and overcome some of the limitations of previous work, thirty native German speakers were recorded in a production study while interacting with a confederate (one of the experimenters) in a referential game, following expermintal procedures typical of the field.
The participants had to verbally instruct the confederate to select a specified target object out of four objects presented on a screen.
The subject and confederate were seated at the opposite sides of a table, each facing one of two computer screens.
The participant and the experimenter could not see each other nor each others' screens.
Figure \@ref(fig:procedural) shows the experimental procedure time-line.
After a familiarization phase, the subject first saw four colored objects in the top left, top right, bottom left, and bottom right corners of the screen.
One of the objects served as the target, another as the competitor, and the remaining two objects served as distractors.
Objects were referred to using noun phrases consisting of an adjective modifier denoting color and a modified object (e.g. *Gelbe Zitrone* 'yellow lemon', *Rote Gurke* 'red cucumber', *Rote Socken* 'red socks').

@@

The two sentence prompts were used to create a focus contrast between the competitor and the target object.
Focused units denote the set of all (contextually relevant) alternatives [e.g. @rooth1992theory].
Concretely, a focus contrast marks one of more elements in a sentence as prominent, by different linguistic means depending on the language.
For instance, if the competitor and target objects differ but their color did not (e.g. *yellow banana* vs *yellow tomato*), the noun is said to be in focus (the Noun Focus condition, NF).
If the objects are the same but differ in color (e.g. *yellow banana* vs *blue banana*), the color adjective is in focus (the Adjective Focus condition, AF).
If both the color and the object differ (e.g. *yellow banana* vs *blue tomato*), the the whole noun phrase is in focus (the Adjective/Noun Focus condition, ANF).
The NF condition constituted the experimentally relevant condition, while the AF and ANF conditions acted as fillers.
Crucially, the color-object combinations in the Noun Focus (NF) condition were manipulated with respect to their typicality.
The combinations were either typical (e.g. *orange mandarin*), medium typical (e.g. *green tomato*), or atypical (e.g. *yellow cherry*), as established by a norming study that was conducted prior to the production experiment just described.^[A detailed description of the norming and production studies from the *Prosodic encoding of redundant referring expressions* project, which was given to the analysts with the data set, can be found in `methods_norm_prod.pdf` at (temporary link) <https://osf.io/5agn9/?view_only=3429acfd8eae4f0aafe2e17ad5b0983f>]
Each subject produced 15 critical trials (NF condition).
Each trial was repeated twice, yielding a total of 30 trials per participant and a grand-total of 900 ($15 \times 2 \times 30$ participants) spoken utterances.

For the present study, N analysis teams will receive access to the entire data set generated by the production study.
The data set is constituded by audio recordings and annotation files in a format that is typical for the field.
The teams will be instructed to answer the following research question, using the provided data set: *Do speakers acoustically modify utterances to signal atypical word combinations?*

>>>>

Speech is a complex signal that is characterized by structurally different acoustic landmarks distributed throughout different temporal domains.
Thus, choosing how to measure a communictive intention of interest is an important analytic step.
Take for example the following sentence in (1).

@@

The multiplicity of phonetic measurements grows exponentially if we look at larger temporal domains as is the case for suprasegmental aspects of speech.
For example, studies investigating acoustic correlates of word stress (e.g. the difference between *ínsight* and *incíte*) use a wide variety of measurements, including temporal characteristics (duration of certain segments or sub-segmental intervals), spectral characteristics (intensity, formants, and spectral tilt), and measurements related to fundamental frequency (f0) [e.g., @gordon2017acoustic].
Moving on to the expression of higher-level communicative functions, like information structure and discourse pragmatics, relevant acoustic cues can be distributed throughout even larger domains, such as phrases and whole utterances [e.g., @ladd2008intonational].
Differences in position, shape, and alignment of pitch modulations over multiple locations within a sentence are correlated with differences in discourse functions [e.g., @niebuhr2011].
The latter can also be expressed by global vs local pitch modulations [@heuven2002], as well as acoustic information within the temporal or spectral domain [e.g., @van2005speech].
Extra-linguistic information, like the speaker's intentions, levels of emotional arousal or social identity, are also conveyed by broad-domain parameters, such as voice quality, rhythm, and pitch [@foulkes2006; @ogden2004; @white2009].

In short, when testing hypotheses on speakers' intentions with speech production data, researchers are faced with many choices and possibilities.
The larger the functional domain (e.g. segments vs words vs utterances), the higher the number of conceivable operationalizations.
For example, several decisions have to be made when comparing the two realizations of the sentence in (1), one of which is intended to signal emphasis on *another* and one of which emphasizes *Zoom* (see 2a and 2b).

@@

Do we compare only the word *another* in (2a) and (2b), or also the word *Zoom*?
Do we measure utterance-wide acoustic profiles, whole words, or just stressed syllables?
Do we average across the chosen time domain or do we measure a specific point in time?
Do we measure fundamental frequency, intensity, or something else [@stevens2000]?

@@

When looking at phrase-level temporal domains, the number of alternative analytic pipelines increases substantially.
Figure \@ref(fig:forkingPaths)A shows a typical example of a decision tree with which speech researchers are often confronted.
Each of the four analytic decisions in the example have different possible options.
Here only one particular path has been taken. A different one would likely produce different results and might lead to different conclusions.
Once we have decided to compare fundamental frequency of the word *another* across the two utterances, there are still many choices to be made, all of which need to be justified.
As Figures \@ref(fig:forkingPaths)B-C illustrate, we could measure f0 at specific points in time like the onset of the temporal window, the offset, or the midpoint.
We could also measure the value or time of the minimum or maximum f0 value.
We could summarize f0 across the entire window and extract the mean, median or standard deviation of f0, all of which have been used to analyze speech data in previous work [see @gordon2017acoustic].
But the journey in the garden of analytic paths goes on.
Other important operationalization steps could involve filtering the audio signal, smoothing the extracted f0 track, removing values that substantially deviate from surrounding values or expectations, either manually or automatically, and so on.

These decisions are intended to be made prior to any statistical analysis, but are at times revised a posteriori in light of unforeseen or surprising outcomes (i.e. after data collection and/or preliminary analyses).
These myriads of possible decisions are multiplied by those researcher degrees of freedom related to statistical analysis [e.g. @wicherts2016].

In sum, speech data is made of complex physical signals that generates an as-of-yet unappreciated amount of analytic flexibility in the choice of measures and operationalizations.
The present paper probes this garden of forking paths in the analysis of speech.
To assess the variability in data analysis pipelines, including both operationalization and statistical analysis, across independent researchers, we provide analytic teams with an experimentally elicited speech production data set.
The data set derives from the unpublished research project *Prosodic encoding of redundant referring expressions*, which set out to investigate whether speakers acoustically modify utterances to signal unexpected referring expressions.^[Results of this research project were neither published nor publicly presented and are stored on a private OSF repository.] 
In the following section we introduce the research question and the experimental procedure of said project, and we describe the resulting data set as used in the current study.

@@

Referring is one of the most basic and prevalent uses of language and one of the most widely researched areas in language science. 
When trying to refer to a banana, what does a speaker say and how do they say it in a given context.
The context within which an entity occurs (i.e., with other non-fruits, other fruits, or other bananas) plays a large part in determining the choice of referring expressions.
Generally, speakers aim to be as informative as possible to uniquely establish reference to the intended object, but they are also resource-efficient in that they avoid redundancy [@grice1975logic].
Thus one would expect the use of a modifier, for example, only if it is necessary for disambiguation.
For instance, one might use the adjective *yellow* to describe a banana in a situation in which there is a yellow and a less ripe green banana available, but not when there is only one banana to begin with.

Despite the coherent idea that speakers are both rational and efficient, there is much evidence that speakers are often over-informative.
Speakers use referring expressions that are more specific than strictly necessary for the unambiguous identification of the intended referent [@sedivy2003pragmatic; @westerbeek2015stored; @rubio2016redundant], which has been argued to facilitate object identification and make communication between speakers and listeners more efficient [@arts2011overspecification; @paraboni2007generating; @rubio2016redundant].
Recent findings suggest that the utility of a referring expression depends on how good it is for a listener (compared to other referring expressions) to identify a target object.
For example, @degen2020redundancy showed that modifiers that are less typical for a given referent (e.g. a blue banana) are more likely to be used in an over-informative scenario (e.g. when there is just one banana).
This account, however, has mainly focused on content selection [@gatt2013we], i.e. what words to use.

Even when morphosyntactically identical expressions are involved, speakers can modulate utterances via suprasegmental acoustic properties like temporal and spectral modifications [e.g., @ladd2008intonational].
Most prominently, languages can use intonation to signal discourse relationships between referents.
Intonation marks discourse-relevant referents for being new or given information, to guide the listeners’ interpretation of incoming messages.
Beyond structuring information relative to the discourse, a few studies suggested that speakers might use intonation to signal atypical lexical combinations [e.g. @dimitrova2008prosodic; @dimitrova2009did].
Referential expressions such as *blue banana* were produced with greater prosodic prominence than more typical referents such as *yellow banana*.
These results are in line with the idea of resource-efficient, rational language users who modulate their speech in order to facilitate listeners' comprehension.
However, the above studies are based on a small sample size (10 participants) and on potentially anti-conservative statistical analyses, leaving reason to doubt the generalizability of the studies' conclusions.

To further illuminate the question of whether speakers modify speech to signal atypical referents, and overcome some of the limitations of previous work, thirty native German speakers were recorded in a production study while interacting with a confederate (one of the experimenters) in a referential game, following experimental procedures typical of the field.
The participants had to verbally instruct the confederate to select a specified target object out of four objects presented on a screen.
The subject and confederate were seated at the opposite sides of a table, each facing one of two computer screens.
The participant and the experimenter could not see each other nor each others' screens.
Figure \@ref(fig:procedural) shows the experimental procedure time-line.
After a familiarization phase, the subject first saw four colored objects in the top left, top right, bottom left, and bottom right corners of the screen.
One of the objects served as the target, another as the competitor, and the remaining two objects served as distractors.
Objects were referred to using noun phrases consisting of an adjective modifier denoting color and a modified object (e.g. *Gelbe Zitrone* 'yellow lemon', *Rote Gurke* 'red cucumber', *Rote Socken* 'red socks').

@@

The two sentence prompts were used to create a focus contrast between the competitor and the target object.
Focused units denote the set of all (contextually relevant) alternatives [e.g. @rooth1992theory].
Concretely, a focus contrast marks one of more elements in a sentence as prominent, by different linguistic means depending on the language.
For instance, if the competitor and target objects differ but their color did not (e.g. *yellow banana* vs *yellow tomato*), the noun is in focus (the Noun Focus condition, NF).
If the objects are the same but differ in color (e.g. *yellow banana* vs *blue banana*), the color adjective is in focus (the Adjective Focus condition, AF).
If both the color and the object differ (e.g. *yellow banana* vs *blue tomato*), the the whole noun phrase is in focus (the Adjective/Noun Focus condition, ANF).
The NF condition constituted the experimentally relevant condition, while the AF and ANF conditions acted as fillers.
Crucially, the color-object combinations in the Noun Focus (NF) condition were manipulated with respect to their typicality.
The combinations were either typical (e.g. *orange mandarin*), medium typical (e.g. *green tomato*), or atypical (e.g. *yellow cherry*), as established by a norming study that was conducted prior to the production experiment just described.^[A detailed description of the norming and production studies from the *Prosodic encoding of redundant referring expressions* project, which was given to the analysts with the data set, can be found in `methods_norm_prod.pdf` at (temporary link) <https://osf.io/5agn9/?view_only=3429acfd8eae4f0aafe2e17ad5b0983f>]
Each subject produced 15 critical trials (NF condition).
Each trial was repeated twice, yielding a total of 30 trials per participant and a grand-total of 900 ($15 \times 2 \times 30$ participants) spoken utterances.

For the present study, N analysis teams will receive access to the entire data set generated by the production study.
The data set is constituted by audio recordings and annotation files in a format that is typical for the field.
The teams will be instructed to answer the following research question, using the provided data set: *Do speakers acoustically modify utterances to signal atypical word combinations?*
